Quantum Hoare Types

### 2020-02-14 ms-paper/ms-paper.tex
Add a line for lift and force
Minor suggested changes incorporated

As programming languages researchers, we think it will be more useful to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start. Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages~\cite{nanevski_hoare_2008}, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.

@@

This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counterintuitive! One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured, the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart). We will see how this can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}. We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.

@@

\section{Running Example}
\textit{Move teleportation example here}

\section{Hoare Types for Quantum Programming Languages}
In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol. We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.

@@

In Fig. \ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right. It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$. The function first entangles the latter two qubits into a Bell state. Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob. Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice. The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}. The specification that the function must meet is written using a Hoare type, which we describe next.

@@

Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts. Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits). $!A$ represents duplicable/reusable data. The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$. $\mathrm{Id_A}(M, N)$ is propositional equality. $\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.

@@

The primitive quantum gates that are treated as constants in the language have types such as follows:
\begin{mathpar}
	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
	\and
	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
	\and
	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
	\and
	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
\end{mathpar}

>>>>

As programming language researchers, we think it will be more useful to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start. Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages~\cite{nanevski_hoare_2008}, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.

@@

This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counterintuitive! One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured, the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart). We will see how this property can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}. We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.

@@

\section{Running Example}

@@

In Fig. \ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right. It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$. The function first entangles the latter two qubits into a Bell state. Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob. Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice. The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}. The specification that the function must meet is written using a Hoare type, which we describe next.
\section{Hoare Types for Quantum Programming Languages}
In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol. We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.

@@

Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts. Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits). $!A$ represents duplicable/reusable data. The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$. $\mathrm{Id_A}(M, N)$ is propositional equality. $\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse. lift and force convert between linear and reusable types.

@@

The primitive quantum gates that are treated as constants in the language have types such as follows:
\begin{mathpar}
	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
	\and
	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
	\\
	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
	\and
	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
\end{mathpar}
